Supplemental Information A: Full Annotated Analysis and Reproducible Code

### 2022-11-27 Supplemental-Material-A-Annotated-Analysis.Rmd
m&r init
:pencil: results intro

@@

## Feature Selection

<div class="alert alert-warning alert-dismissible fade in" role="alert">
  <a class="close" data-dismiss="alert" aria-label="close">&times;</a>
  <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/>
  Here we might have to make a selection of the features we extract because some of the clustering algorithms might not be able to deal with the many different features. See below
</div>

Feature selection types:

-   manual feature and variable selection
-   wrapper: procedures that add and/or remove predictors to find the optimal combination that maximizes model performance.
-   filter: relevance of the predictors outside of the predictive models and subsequently model only the predictors that pass some criterion.
-   dimension reduction for clustering. No selection more compression of redundant variance.

```{r manual feature selection}

```{r automated feature selection}

@@

### Main Anlaysis: Across Studies

The main clustering procedure, uses the features calculated for the relevant variables across all three studies.

#### Centroid-based

For our main method, we use the centroid-based k-means clustering. The procedure needs the researcher to make a selection on the optimal number of clusters in order to run the main algorithm. We use the Elbow and the Silhouette method, which both suggest two clusters to be the optimal solution.

@@

#### Distribution-based

For the distribution-based clustering we use the commonly used Gaussian mixture modeling (GMM) procedure. This procedure assumes  independent Gaussian distributions for each each cluster. These distributions are substantially better at identifying clusters if the data comes from multivariate Gaussian distributions (i.e., ellipses in a two dimensional space).

We again begin by determining the optimal number of clusters. For GMMs this is done using common information criteria, such as the BIC.

<div class="alert alert-warning alert-dismissible fade in" role="alert">
  <a class="close" data-dismiss="alert" aria-label="close">&times;</a>
  <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/>
  This is where we run into trouble. These plots are relatively uncommon and according to the internet this is most likely a problem of too many (non-distinguishing or highly correlated) features. A good paper on this is <a href="https://proceedings.neurips.cc/paper/2002/file/e58aea67b01fa747687f038dfde066f6-Paper.pdf">Law et al. (2002)</a>.
</div>

```{r gaussian mixture model}
library(mclust)

scaled_data <- featFullImp$featuresImpZMat %>% select(-ends_with(featureExclusion))

# scaled_data <- featS1Imp$featuresImpZMat  %>% #select(starts_with(varNamS123Int))
#   select(-any_of(names(which(colSums(is.na(.)) > 0))))

BIC <- mclustBIC(scaled_data)
plot(BIC)
summary(BIC)
mod1 <- Mclust(scaled_data)
summary(mod1, parameters = FALSE)

# multivariate <- densityMclust(scaled_data, plot=FALSE)
# summary(multivariate)            
# plot(multivariate, what = "BIC")

# merge(featOutZRowNam, as.data.frame(mod1$classification), by="row.names") %>% View

#table(class, mod1$classification)

#mod1dr <- MclustDR(mod1)
#summary(mod1dr, parameters = FALSE)
#plot(mod1dr, what = "classification")
```

<div class="alert alert-warning alert-dismissible fade in" role="alert">
  <a class="close" data-dismiss="alert" aria-label="close">&times;</a>
  <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/>
  The feature selection issue is further highlighted when we do the GMMs by feature. Where we find a reasonable number of clusters for each feature. 
</div>

```{r MClust feature specific cluster numbers}
scaled_data <- featFullImp$featuresImpZMat %>% select(-ends_with(featureExclusion))
nMClust <- list()
for(f in unique(gsub("^.*?\\_", "", names(scaled_data)))){
  scaled_data <- featFullImp$featuresImpZMat %>% select(ends_with(f))
  mod1 <- Mclust(scaled_data)
  #summary(mod1, parameters = FALSE)
  nMClust[[f]] <- mod1$G
}

nMClust %>%
  unlist %>%
  data.frame(feature=names(.), G=., row.names=NULL) %>%
  kbl(.,
      escape = FALSE,
      booktabs = T,
      align = c("l", "c"),
      digits=2,
      col.names = c(
        "Feature",
        "Number of Clusters"
      ),
      caption = "Number of MClust Clusters if separated by Feature") %>%
  kable_classic(
    full_width = F,
    lightable_options = "hover",
    html_font = "Cambria"
  )

#### Density-based

For the density-based cluster analysis we focus on the *DBSCAN* algorithm ('Density-Based Spatial Clustering of Applications with Noise') as well as its extension the *OPTICS* procedure ('Ordering Points To Identify the Clustering Structure'). This procedure is more bottom up than the previous two approaches. Instead of providing the number of clusters, a cluster is specified if a minimum number of other points (the MinPts parameter) are within a given radius (the epsilon parameter). Together these two parameters determine a density at which a group of data points are considered a cluster. While this description is grossly oversimplified, it highlights one of the advantages of the procedures, namely that dense regions can expand to be any shape in the d-dimensional space.

<div class="alert alert-warning alert-dismissible fade in" role="alert">
  <a class="close" data-dismiss="alert" aria-label="close">&times;</a>
  <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/>
  Same feature selection issue as above. 
</div>

```{r dbscan}
library(dbscan)
library(fpc)

#scaled_data <- featFullImp$featuresImpZMat %>% select(starts_with(varNamS123Int))
scaled_data <- featFullImp$featuresImpZMat %>% select(-ends_with(featureExclusion))  # %>% select(starts_with(varNamS123NoInt))#%>%
  #select(-any_of(names(which(colSums(is.na(scaled_data)) > 0))))

#is.na(scaled_data) %>% table

# to plot the eps values
eps_plot = kNNdistplot(scaled_data, k=3)
# to draw an optimum line
eps_plot %>% abline(h = 11, lty = 2)

potentialMinPts <- round(SciViews::ln(nrow(scaled_data)))
eps <- 11
minPts <- 2

fviz_cluster(res.fpc, scaled_data, geom = "point", ggtheme = theme_Publication())
#hullplot(scaled_data, res.fpc)

<div class="alert alert-warning alert-dismissible fade in" role="alert">
  <a class="close" data-dismiss="alert" aria-label="close">&times;</a>
  <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/>
  And again the number of clusters is reasonable if we consider the features separately: 
</div>

```{r dbscan feature specific cluister numbers}
scaled_data <- featFullImp$featuresImpZMat %>% select(-ends_with(featureExclusion))
minPts <- 3
xi <- 0.04 #0.05
nDbscan <- list()
for(f in unique(gsub("^.*?\\_", "", names(scaled_data)))){
  scaled_data <- featFullImp$featuresImpZMat %>% select(ends_with(f))
  #res.fpc <- dbscan::dbscan(scaled_data, eps=eps_cl, MinPts=minPts)

  res.optics <- dbscan::optics(scaled_data, minPts = 3)

  #plot(res.optics, col = "grey")
  #dend <- as.dendrogram(res.optics)
  #plot(dend, ylab = "Reachability dist.", leaflab = "none")
  #extractDBSCAN(res.optics, eps_cl = 19.12)

  res <- extractXi(res.optics, xi = xi)
  res$clusters_xi
  #plot(res)
  #hullplot(scaled_data, res)

  #summary(res.optics, parameters = FALSE)
  nDbscan[[f]] <- nrow(res$clusters_xi)
}

nDbscan %>%
  unlist %>%
  data.frame(feature=names(.), G=., row.names=NULL) %>%
  kbl(.,
      escape = FALSE,
      booktabs = T,
      align = c("l", "c"),
      digits=2,
      col.names = c(
        "Feature",
        "Number of Clusters"
      ),
      caption = "Number of OPTICS Clusters if separated by Feature") %>%
  kable_classic(
    full_width = F,
    lightable_options = "hover",
    html_font = "Cambria"
  )
```

### Follow-up: By Response Type

_**currently removed for faster render until feature selection issue is resolved**_

### Follow-up: By Study

_**currently removed for faster render until feature selection issue is resolved**_

>>>>

```{r export cleaned and reduced datasets}
# save cleaned 
save(dtS1, dtS2, dtS3, file = "data/s123Full.RData")

# save reduced
save(list = ls(pattern = "Red$"), file = "data/s123Red.RData")
```

@@

## Feature Reduction

To avoid the dimensionality curse we reduce the extracted features using a principal component analysis.

MORE INFO NEEDED

@@

The main clustering procedure, uses the features calculated for the relevant variables across all three studies. For our main method, we use the centroid-based k-means clustering. The procedure needs the researcher to make a selection on the optimal number of clusters in order to run the main algorithm. We use the Elbow and the Silhouette method, which both suggest two clusters to be the optimal solution.

@@

<!-- #### Distribution-based -->

<!-- For the distribution-based clustering we use the commonly used Gaussian mixture modeling (GMM) procedure. This procedure assumes  independent Gaussian distributions for each each cluster. These distributions are substantially better at identifying clusters if the data comes from multivariate Gaussian distributions (i.e., ellipses in a two dimensional space).  -->

<!-- We again begin by determining the optimal number of clusters. For GMMs this is done using common information criteria, such as the BIC. -->

<!-- <div class="alert alert-warning alert-dismissible fade in" role="alert"> -->
<!--   <a class="close" data-dismiss="alert" aria-label="close">&times;</a> -->
<!--   <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/> -->
<!--   This is where we run into trouble. These plots are relatively uncommon and according to the internet this is most likely a problem of too many (non-distinguishing or highly correlated) features. A good paper on this is <a href="https://proceedings.neurips.cc/paper/2002/file/e58aea67b01fa747687f038dfde066f6-Paper.pdf">Law et al. (2002)</a>. -->
<!-- </div> -->

<!-- ```{r gaussian mixture model} -->
<!-- library(mclust) -->

<!-- scaled_data <- featFullImp$featuresImpZMat %>% select(-ends_with(featureExclusion)) -->

<!-- # scaled_data <- featS1Imp$featuresImpZMat  %>% #select(starts_with(varNamS123Int)) -->
<!-- #   select(-any_of(names(which(colSums(is.na(.)) > 0)))) -->

<!-- BIC <- mclustBIC(scaled_data) -->
<!-- plot(BIC) -->
<!-- summary(BIC) -->
<!-- mod1 <- Mclust(scaled_data) -->
<!-- summary(mod1, parameters = FALSE) -->

<!-- # multivariate <- densityMclust(scaled_data, plot=FALSE) -->
<!-- # summary(multivariate)             -->
<!-- # plot(multivariate, what = "BIC") -->

<!-- # merge(featOutZRowNam, as.data.frame(mod1$classification), by="row.names") %>% View -->

<!-- #table(class, mod1$classification) -->

<!-- #mod1dr <- MclustDR(mod1) -->
<!-- #summary(mod1dr, parameters = FALSE) -->
<!-- #plot(mod1dr, what = "classification") -->
<!-- ``` -->

<!-- <div class="alert alert-warning alert-dismissible fade in" role="alert"> -->
<!--   <a class="close" data-dismiss="alert" aria-label="close">&times;</a> -->
<!--   <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/> -->
<!--   The feature selection issue is further highlighted when we do the GMMs by feature. Where we find a reasonable number of clusters for each feature.  -->
<!-- </div> -->

<!-- ```{r MClust feature specific cluster numbers} -->
<!-- scaled_data <- featFullImp$featuresImpZMat %>% select(-ends_with(featureExclusion)) -->
<!-- nMClust <- list() -->
<!-- for(f in unique(gsub("^.*?\\_", "", names(scaled_data)))){ -->
<!--   scaled_data <- featFullImp$featuresImpZMat %>% select(ends_with(f)) -->
<!--   mod1 <- Mclust(scaled_data) -->
<!--   #summary(mod1, parameters = FALSE) -->
<!--   nMClust[[f]] <- mod1$G -->
<!-- } -->

<!-- nMClust %>% -->
<!--   unlist %>% -->
<!--   data.frame(feature=names(.), G=., row.names=NULL) %>% -->
<!--   kbl(., -->
<!--       escape = FALSE, -->
<!--       booktabs = T, -->
<!--       align = c("l", "c"), -->
<!--       digits=2, -->
<!--       col.names = c( -->
<!--         "Feature", -->
<!--         "Number of Clusters" -->
<!--       ), -->
<!--       caption = "Number of MClust Clusters if separated by Feature") %>% -->
<!--   kable_classic( -->
<!--     full_width = F, -->
<!--     lightable_options = "hover", -->
<!--     html_font = "Cambria" -->
<!--   ) -->

<!-- ``` -->

<!-- #### Density-based -->

<!-- For the density-based cluster analysis we focus on the *DBSCAN* algorithm ('Density-Based Spatial Clustering of Applications with Noise') as well as its extension the *OPTICS* procedure ('Ordering Points To Identify the Clustering Structure'). This procedure is more bottom up than the previous two approaches. Instead of providing the number of clusters, a cluster is specified if a minimum number of other points (the MinPts parameter) are within a given radius (the epsilon parameter). Together these two parameters determine a density at which a group of data points are considered a cluster. While this description is grossly oversimplified, it highlights one of the advantages of the procedures, namely that dense regions can expand to be any shape in the d-dimensional space. -->

<!-- <div class="alert alert-warning alert-dismissible fade in" role="alert"> -->
<!--   <a class="close" data-dismiss="alert" aria-label="close">&times;</a> -->
<!--   <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/> -->
<!--   Same feature selection issue as above.  -->
<!-- </div> -->

<!-- ```{r dbscan} -->
<!-- library(dbscan) -->
<!-- library(fpc) -->

<!-- #scaled_data <- featFullImp$featuresImpZMat %>% select(starts_with(varNamS123Int)) -->
<!-- scaled_data <- featFullImp$featuresImpZMat %>% select(-ends_with(featureExclusion))  # %>% select(starts_with(varNamS123NoInt))#%>% -->
<!--   #select(-any_of(names(which(colSums(is.na(scaled_data)) > 0)))) -->

<!-- #is.na(scaled_data) %>% table -->

<!-- # to plot the eps values -->
<!-- eps_plot = kNNdistplot(scaled_data, k=3) -->
<!-- # to draw an optimum line -->
<!-- eps_plot %>% abline(h = 11, lty = 2) -->

<!-- potentialMinPts <- round(SciViews::ln(nrow(scaled_data))) -->
<!-- eps <- 11 -->
<!-- minPts <- 2 -->

<!-- res.fpc <- fpc::dbscan(scaled_data, eps=eps, MinPts=minPts, scale=FALSE, showplot=0) -->
<!-- res.db <- dbscan::dbscan(scaled_data, eps=eps, minPts=minPts) -->

<!-- resEqual.check <- all(res.fpc$cluster == res.db$cluster) -->
<!-- nClustDBSCAN <- length(unique(res.fpc$cluster))-1 -->

<!-- fviz_cluster(res.fpc, scaled_data, geom = "point", ggtheme = theme_Publication()) -->
<!-- #hullplot(scaled_data, res.fpc) -->

<!-- ``` -->

<!-- <div class="alert alert-warning alert-dismissible fade in" role="alert"> -->
<!--   <a class="close" data-dismiss="alert" aria-label="close">&times;</a> -->
<!--   <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/> -->
<!--   And again the number of clusters is reasonable if we consider the features separately:  -->
<!-- </div> -->

<!-- ```{r dbscan feature specific cluister numbers} -->
<!-- scaled_data <- featFullImp$featuresImpZMat %>% select(-ends_with(featureExclusion)) -->
<!-- minPts <- 3 -->
<!-- xi <- 0.04 #0.05 -->
<!-- nDbscan <- list() -->
<!-- for(f in unique(gsub("^.*?\\_", "", names(scaled_data)))){ -->
<!--   scaled_data <- featFullImp$featuresImpZMat %>% select(ends_with(f)) -->
<!--   #res.fpc <- dbscan::dbscan(scaled_data, eps=eps_cl, MinPts=minPts) -->

<!--   res.optics <- dbscan::optics(scaled_data, minPts = 3) -->

<!--   #plot(res.optics, col = "grey") -->
<!--   #dend <- as.dendrogram(res.optics) -->
<!--   #plot(dend, ylab = "Reachability dist.", leaflab = "none") -->
<!--   #extractDBSCAN(res.optics, eps_cl = 19.12) -->

<!--   res <- extractXi(res.optics, xi = xi) -->
<!--   res$clusters_xi -->
<!--   #plot(res) -->
<!--   #hullplot(scaled_data, res) -->

<!--   #summary(res.optics, parameters = FALSE) -->
<!--   nDbscan[[f]] <- nrow(res$clusters_xi) -->
<!-- } -->

<!-- nDbscan %>% -->
<!--   unlist %>% -->
<!--   data.frame(feature=names(.), G=., row.names=NULL) %>% -->
<!--   kbl(., -->
<!--       escape = FALSE, -->
<!--       booktabs = T, -->
<!--       align = c("l", "c"), -->
<!--       digits=2, -->
<!--       col.names = c( -->
<!--         "Feature", -->
<!--         "Number of Clusters" -->
<!--       ), -->
<!--       caption = "Number of OPTICS Clusters if separated by Feature") %>% -->
<!--   kable_classic( -->
<!--     full_width = F, -->
<!--     lightable_options = "hover", -->
<!--     html_font = "Cambria" -->
<!--   ) -->
<!-- ``` -->

<!-- ### Follow-up: By Response Type -->
<!-- Potential future project -->

<!-- ### Follow-up: By Study -->
<!-- Potential future project -->

### 2022-11-27 methods-and-results.Rmd
:pencil: results intro

# Sample Descriptives Table --- wider version
allMlCor %>%
  t %>%
  kbl(.,
      format = "latex",
      caption = "Correlation Table and Descriptive Statistics",
      booktabs = TRUE,
      align = c("l", rep("c", ncol(.) - 1))) %>%
  kable_classic() %>%
  add_header_above(
    .,
    c(
      "",
      "Correlations" = ncol(allMlCor),
      "Descriptives" = nrow(allMlCor) - ncol(allMlCor)
    )
  ) %>%
  footnote(
    general = c(
      "Upper triangle: Between-person correlations;",
      "Lower triangle: Within-person correlations;",
      "*** p < .001, ** p < .01,  * p < .05"
    )
  ) %>%
  kable_styling(latex_options = "scale_down") %>%
  gsub("\\begin{table}", "\\begin{sidewaystable}\n\\centering", ., fixed = TRUE) %>%
  gsub("\\end{table}", "\\end{sidewaystable}", ., fixed = TRUE) %>%
  save_kable("Tables/descrWide.tex")

@@

\subsection{Analysis and Results}

\subsubsection{Feature extraction}

>>>>

# Sample Descriptives Table --- wider version
allMlCor %>%
  t %>%
  as.data.frame %>%
  kbl(.,
      format = "latex",
      caption = "Correlation Table and Descriptive Statistics",
      booktabs = TRUE,
      align = c("l", rep("c", ncol(.) - 1))) %>%
  kable_classic() %>%
  add_header_above(
    .,
    c(
      "",
      "Correlations" = ncol(allMlCor),
      "Descriptives" = nrow(allMlCor) - ncol(allMlCor)
    )
  ) %>%
  footnote(
    general = c(
      "Upper triangle: Between-person correlations;",
      "Lower triangle: Within-person correlations;",
      "*** p < .001, ** p < .01,  * p < .05"
    )
  ) %>%
  kable_styling(latex_options = "scale_down") %>%
  gsub("\\begin{table}", "\\begin{sidewaystable}\n\\centering", ., fixed = TRUE) %>%
  gsub("\\end{table}", "\\end{sidewaystable}", ., fixed = TRUE) %>%
  save_kable("Tables/descrWide.tex")

@@

\subsection{Analysis and Results}
To perform the main analysis on our selected variables, we follow the three extract--reduce--cluster steps in sequential order. For each of the steps fully reproducible and annotated code is available in Online Supplemental A as well as in the accompanying OSF- and GitHub repositories (REFERENCES).

\subsubsection{Feature extraction}
```{r featMiss, include=FALSE}
pMissFull <- pMissFeat(
  all = featFull$features,
  contact = featFullContact$features,
  nocontact = featFullNoContact$features,
  title = "Feature-wise Missingess Across all Studies"
)
```

For the features, we selected and extracted one of each of the features proposed in \tblref{tab:esmFeatures}. We particularly chose the more robust median and median absolute deviation (MAD) for central tendency and variability. Within the variability structure, we chose the slightly more intuitive mean absolute change (MAC) for stability and stayed with the common lag--1 autocorrelation for inertia. For the trend summaries, we chose the overall linear regression slope to describe the linear trend and we summarized the nonlinear trend with the estimated degrees of freedom of an empty GAM spline model (edf) --- the edf summarizes the \textit{wiggliness} of the spline trend line.

We include a function that automatically extracts and prepares a large selection of the time series feature operationalizations presented in \tblref{tab:esmFeatures} in our GitHub repository (see the \texttt{featureExtractor} function).

For the full dataset across the three studies, we find that about `r format(round(pMissFull$pmiss %>% filter(set == "All") %>% select(pMissTotal) %>% pull, 2), nsmall=2)`\% of the extracted features are missing across the `r ncol(featFull$features)` features per participant. These missing features are imputed with a single predictive mean matching imputation using the MICE library.

### 2022-11-27 methods-and-results.tex
:pencil: results intro

@@

>>>>

To perform the main analysis on our selected variables, we follow the
three extract--reduce--cluster steps in sequential order. For each of
the steps fully reproducible and annotated code is available in Online
Supplemental A as well as in the accompanying OSF- and GitHub
repositories (REFERENCES). This includes a function that automatically
extracts and prepares all of the time series features presented in
\tblref{tab:esmFeatures} (see the \texttt{featureExtractor} function in
our GitHub repository).

@@

For the

For the full dataset across the three studies, we find that about 1.66\%
of the extracted features are missing across the 145 features per
participant. These missing features are imputed with a single predictive
mean matching imputation using the MICE library.
